---
title: "Poisson"
output: html_document
---

Suppose $X$ is Poi($\lambda$).
If we see $X=x$, how good is the normal approximation to the likelihood for l=log(lambda) ?

Here for x>0 we approximate poisson likelihood for l=log(lambda) by
by normal, with mean log(x) and variance 1/x.

When x=0 I chose the mean to be log(0.01) and the variance so that the approximation is accurate in range l=log(0.01) to log(1) [More precisely ratios are accurate in this range, and particularly the likelihood ratio for rate 0.01 vs 1 is "accurate"]. Could of course change from 0.01 when very small rates are of interest....

```{r}
#
plot_poisson_loglik=function(x,range1=1,range2=1,...){
  m = ifelse(x==0,log(0.01),log(x))
  sd_approx = ifelse(x==0,-m/sqrt(2),exp(m)^-0.5)
  l = m + seq(-range1,range2,length=100)
  loglik = dpois(x,exp(l),log=TRUE)
  plot(l, loglik-max(loglik),type="l",...)
  abline(v=log(x))
  normlik = log(dnorm(l,m,sd=sd_approx))
  lines(l,normlik-max(normlik),col="red")
}
plot_poisson_loglik(0,main="x=0",range1=1,range2=5)
plot_poisson_loglik(1,main="x=1")
plot_poisson_loglik(2,main="x=2")
plot_poisson_loglik(100,main="x=100")
```

It looks, visually, like approximation is pretty good for even $x$ as small as $1$,
but the zero case is the one that really is problematic.

Notice that the $y$ axis for $x=100$ is 100 times bigger than $x=1$. This
is easy to see why mathematically.

Let $X \sim Poi(\lambda)$ and $l = \log(\lambda)$. Suppose we observe $X=x$.
Let $l_0 = \log(x)$, which is the maximum likelihood estimate for $l$.

Then the log likelihood ratio for $l$ vs $l_0$ is: 
$$logLR = l(l) - l(l_0) = x (l-l_0) - (exp(l) - exp(l_0))$$
$$ = x(\delta - exp(\delta) + 1)$$
where $\delta = l-l_0$, and we have substituted $x=exp(l_0)$.


Could probably improve $x=1,2$ cases by using a mixture of normals.

Compare with usual approximation, X is N(lambda, lambda). CHECK THIS!
```{r}
plot_poisson_loglik2=function(x,range1=1,range2=1,...){
  m = ifelse(x==0,0,x)
  l = log(m) + seq(-range1,range2,length=100)
  loglik = dpois(x,exp(l),log=TRUE)
  plot(l, loglik-max(loglik),type="l",...)
  abline(v=log(x))
  normlik = log(dnorm(exp(l),m,sd=exp(l)))
  lines(l,normlik-max(normlik),col="red")
}
plot_poisson_loglik2(1,main="x=1")
plot_poisson_loglik2(100,main="x=100")
```


```{r}
n=20
set.seed(1)
x = rpois(n,1)
y = ifelse(x==0,0.01,x)
m = (y-1)/(2*y)   #this was an error, but seems to work curiously well...
sd_approx = sqrt(1/(y)) 

m = (y-1)
sd_approx = rep(1,length(m))
#m=ifelse(x==0,log(0.01),log(x))
#sd_approx = ifelse(x==0,-m/sqrt(2),exp(m)^-0.5)
l = seq(-1,1,length=100)
loglik.pois = loglik.norm = matrix(0,nrow=n,ncol=100)
for(i in 1:n){loglik.pois[i,] = dpois(x[i],exp(l),log=TRUE)}
for(i in 1:n){loglik.norm[i,] = dnorm(m[i],l,sd=sd_approx[i],log=TRUE)}
loglik.pois.tot = apply(loglik.pois,2,sum)
loglik.norm.tot = apply(loglik.norm,2,sum)
plot(l,loglik.pois.tot-max(loglik.pois.tot))
lines(l,loglik.norm.tot-max(loglik.norm.tot),col=2)

```


```{r}
n=200
set.seed(1)
x = rpois(n,10)

lambda0  = mean(x)
m = log(lambda0) + lambda0*(x-lambda0)
sd_approx = rep(sqrt(1/lambda0),n)
l = log(lambda0) + seq(-1,1,length=100)
loglik.pois = loglik.norm = matrix(0,nrow=n,ncol=100)
for(i in 1:n){loglik.pois[i,] = dpois(x[i],exp(l),log=TRUE)}
for(i in 1:n){loglik.norm[i,] = dnorm(m[i],l,sd=sd_approx[i],log=TRUE)}
loglik.pois.tot = apply(loglik.pois,2,sum)
loglik.norm.tot = apply(loglik.norm,2,sum)
plot(l,loglik.pois.tot-max(loglik.pois.tot))
lines(l,loglik.norm.tot-max(loglik.norm.tot),col=2)

```

